---
title: "Movie Recommender Capstone"
author: "KJ Monahan"
date: "2026-03-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The goal of this project was to create a machine-learning model that can predict
how a given user will rate a given movie on a scale of 0.5 stars to 5 stars. These
predictions could then be used to recommend new movies to a user.

The dataset contains 10 million ratings of movies by users, as well as the rating
timestamp, year released, and genre(s) of the movie. Since users rate varying
numbers of movies and movies vary in the number of people who have watched and
rated them, the matrix of users and movies is a sparse matrix.

The final model uses a combination of bias effects from known data (user, movie, 
and genre) and latent factors.

## Methods/Analysis

### Data Cleaning
Initially, the data contained the movie release year as part of the movie title.
After creating my training set and test set (separate from the final holdout test
set to use at the end), I cleaned the data by extracting the year into a 
separate column. My initial hypothesis was that the decade in which the movie was 
released might have an effect on the outcome, since certain eras of cinema 
seemed to be more highly regarded than others. So I used the release year to 
calculate the decade. I also converted the timestamp from a string to a datetime 
and the genres into factors. I did not end up using all of this data in the 
model, but I wasn't sure at this point what I would or would not need.

```{r setup-data, include=FALSE}
library("tidyverse")
library(data.table)
load("rdas/train_set_clean.RData")
load("rdas/test_set_clean.RData")
load("rdas/final_holdout_test_clean.RData")

clamp <- function(x, lower = 0.5, upper = 5)
  pmax(pmin(x, upper), lower)

rmse <- function(r)
  sqrt(mean(r^2))
```

### Analysis
I knew from previous work that there would likely be a user effect and a movie
effect. To verify this, I grouped ratings by each of those factors and reviewed
the results.

```{r users-effect, warning=FALSE}
fit_users <- train_set |> 
  group_by(userId) |> 
  summarize(b_u = mean(rating))
qplot(fit_users$b_u, bins = 10, color = I("black"))
```

```{r movies-effect, warning=FALSE}
fit_movies <- train_set |> 
  group_by(movieId) |> 
  summarise(b_i = mean(rating))
qplot(fit_movies$b_i, bins = 10, color = I("black"))
```
Ratings did appear to vary by user and by movie.

I expected that ratings would vary by genre, and they did seem to.
```{r genres-effect, warning=FALSE}
fit_genres <- train_set |> 
  group_by(genres) |> 
  summarise(b_g = mean(rating))
qplot(fit_genres$b_g, bins = 10, color = I("black"))
```
The shape looks a lot like the movie effect, so it is possible that the 
difference in rating between movies already accounts for genre. However, I 
thought it still might be useful. I could imagine that knowing that, for 
instance, "Drama" movies are higher-rated might help give a more accurate 
prediction for a relatively new drama that didn't have many ratings for the 
individual movie yet.

As discussed earlier, I also looked at whether ratings varied by the decade in
which the movie was released.
```{r decade-effect, warning=FALSE}
fit_decades <- train_set |>
  group_by(decade) |>
  summarise(b_d = mean(rating))
qplot(fit_decades$b_d, bins = 10, color = I("black"))
```
Decades have an interesting shape, clustered between two edges with nothing in the middle.
What the graph reveals (although I didn't realize it until later), however, is 
that the ratings by decade are between 3.4 and a little under 4.0 -- a much 
smaller range than the other factors. The average rating is about 3.5, so while
the average rating for a decade can vary from that a little, overall, ratings 
by decade do not vary enough to make a meaningful impact on the prediction

### Baseline: Average Rating
I first calculated the simplest possible model: assuming the average rating for all ratings,
so that I would have a starting point to verify that I was improving the accuracy. This
proved useful getting any RMSE higher than the baseline was an immediate signal
that I had made a mistake in the model and needed to rework it.

```{r baseline}
mu <- mean(train_set$rating, na.rm = TRUE)
rmse(test_set$rating - mu)
```
### Model 1: Known Effects
I first created a model using the effects of user, movie, and genre. After getting
the overall function in place, I created a grid of tuning parameters to find the 
combination of lambda penalties that gave me the lowest RMSE.

One issue I encountered here was getting the right scale for my lambdas. I 
initially tried finding a value that was multiplied by the total number of rows.
I tested the range 10^-3 through 10^-6. On a training set of roughly
7 million rows, this equated to 7 through 7000. This might have worked, except
that I was failing to also account for the different sizes of groups for each
factor. As a result, I was either under-regularizing genre effects or so heavily 
regularizing user effects that their impact was lost. I switched to using 
constants because I found that easier to comprehend and then used the mean and
median number of ratings per group for each factor to find a more reasonable
scale to use for each.
```{r model-1-known_effects}
fit_als_with_known_effects <- function(data = train_set,
                                       lambda_u = 50,
                                       lambda_m = 100,
                                       lambda_g = 10000,
                                       tol = 1e-6,
                                       max_iter = 500) {
  dt <- as.data.table(copy(data))
  
  N <- nrow(dt)
  mu <- mean(dt$rating)
  
  dt[, resid := rating - mu]
  b_i_dt <- b_u_dt <- b_g_dt <- b_d_dt <- NULL
  prev_loss <- rmse(dt$resid)
  
  for (iter in 1:max_iter) {
    b_i_dt <- dt[, .(b = sum(resid) / (.N + lambda_m)), by = movieId]
    dt[b_i_dt, on = "movieId", bi := i.b]
    dt[, resid := rating - mu - bi]
    
    b_u_dt <- dt[, .(a = sum(resid) / (.N + lambda_u)), by = userId]
    dt[b_u_dt, on = "userId", bu := i.a]
    dt[, resid := rating - mu - bi - bu]
    
    b_g_dt <- dt[, .(c = sum(resid) / (.N + lambda_g)), by = genres]
    dt[b_g_dt, on = "genres", bg := i.c]
    dt[, resid := rating - mu - bi - bu - bg]
    
    loss <- rmse(dt$resid)
    delta <- abs(loss - prev_loss) / prev_loss

    if (delta < tol | delta > 1e07)
      break
    prev_loss <- loss
  }
  
  list(
    mu = mu,
    b_i = as.data.frame(b_i_dt),
    b_u = as.data.frame(b_u_dt),
    b_g = as.data.frame(b_g_dt)
  )
}

fit <- fit_als_with_known_effects(train_set)
mu <- fit$mu
b_i <- setNames(fit$b_i$b, fit$b_i$movieId)
b_u <- setNames(fit$b_u$a, fit$b_u$userId)
b_g <- setNames(fit$b_g$c, fit$b_g$genres)
 
resid <- with(test_set, rating - clamp(mu + 
                                         b_i[as.character(movieId)] + 
                                         b_u[as.character(userId)] + 
                                         b_g[genres]))
rmse(resid)
```

This brought my RMSE down quite a bit, but not yet below 0.86.

### Latent Factors
I knew not just from the course materials but also from general domain
knowledge that there is more that goes into how people rate movies than is
captured in the table columns. My two sisters judge movies based on costuming
and whether the cinematography  is "pretty". I'll bail if the world-building
doesn't make sense. I can't build all of these specific preferences into a model,
but it seemed like including latent factors in the model might account for that.

This proved to be the case. While the first part of the final model is identical
to the previous version, adding latent factors improved the accuracy of the predictions
enough that I could move on to testing it with the final holdout test set.

```{r final-model}
fit_als_with_latent <- function(data = train_set,
                                K = 24,
                                lambda_u = 50,
                                lambda_m = 100,
                                lambda_g = 10000,
                                lambda_pq = 5,
                                tol = 1e-6,
                                max_iter = 500) {
  dt <- as.data.table(copy(data))
  
  N <- nrow(dt)
  mu <- mean(dt$rating)
  
  dt[, resid := rating - mu]
  b_i_dt <- b_u_dt <- b_g_dt <- b_d_dt <- NULL
  prev_loss <- rmse(dt$resid)
  
  for (iter in 1:max_iter) {
    b_i_dt <- dt[, .(b = sum(resid) / (.N + lambda_m)), by = movieId]
    dt[b_i_dt, on = "movieId", bi := i.b]
    dt[, resid := rating - mu - bi]
    
    b_u_dt <- dt[, .(a = sum(resid) / (.N + lambda_u)), by = userId]
    dt[b_u_dt, on = "userId", bu := i.a]
    dt[, resid := rating - mu - bi - bu]
    
    b_g_dt <- dt[, .(c = sum(resid) / (.N + lambda_g)), by = genres]
    dt[b_g_dt, on = "genres", bg := i.c]
    dt[, resid := rating - mu - bi - bu - bg]
    
    loss <- rmse(dt$resid)
    delta <- abs(loss - prev_loss) / prev_loss
    
    if (delta < tol | delta > 1e07)
      break
    prev_loss <- loss
  }
  
  prev_loss <- rmse(dt$resid)
  
  unique_users <- unique(dt$userId)
  unique_movies <- unique(dt$movieId)
  n_users  <- length(unique_users)
  n_movies <- length(unique_movies)
  
  user_int  <- match(dt$userId, unique_users)
  movie_int <- match(dt$movieId, unique_movies)
  resids <- dt$resid
  
  order_by_user <- order(user_int)
  users_by_userId <- user_int[order_by_user]
  movies_by_userId <- movie_int[order_by_user]
  resids_by_userId <- resids[order_by_user]
  
  order_by_movie <- order(movie_int)
  users_by_movieId <- user_int[order_by_movie]
  movies_by_movieId <- movie_int[order_by_movie]
  resids_by_movieId <- resids[order_by_movie]
  
  u_rle <- rle(users_by_userId)
  user_endpoints <- cumsum(u_rle$lengths)
  user_starting_points <- c(1L, user_endpoints[-length(user_endpoints)] + 1L)
  m_rle <- rle(movies_by_movieId)
  movie_endpoints <- cumsum(m_rle$lengths)
  movie_starting_points <- c(1L, movie_endpoints[-length(movie_endpoints)] + 1L)
  
  p <- matrix(rnorm(n_users  * K, mean = 0, sd = 0.01),
              nrow = n_users,
              ncol = K)
  q <- matrix(rnorm(n_movies * K, mean = 0, sd = 0.01),
              nrow = n_movies,
              ncol = K)
  
  lambda_pq_diag <- lambda_pq * diag(K)
  
  for (iter in 1:max_iter) {
    for (i in 1:n_users) {
      current_user <- user_starting_points[i]:user_endpoints[i]
      q_i <- q[movies_by_userId[current_user], , drop = FALSE]
      resids_i <- resids_by_userId[current_user]
      qi_by_qi <- crossprod(q_i) + lambda_pq_diag
      x <- crossprod(q_i, resids_i)
      p[u_rle$values[i], ] <- solve(qi_by_qi, x)
    }
    for (j in 1:n_movies) {
      current_movie <- movie_starting_points[j]:movie_endpoints[j]
      p_j <- p[users_by_movieId[current_movie], , drop = FALSE]
      resids_j <- resids_by_movieId[current_movie]
      pj_by_pj <- crossprod(p_j) + lambda_pq_diag
      x <- crossprod(p_j, resids_j)
      q[m_rle$values[j], ] <- solve(pj_by_pj, x)
    }
    pq <- rowSums(p[user_int, , drop = FALSE] * q[movie_int, , drop = FALSE])
    
    loss <- rmse(resids - pq)
    delta <- abs(loss - prev_loss) / prev_loss
    
    if (delta < tol | delta > 1e07)
      break
    prev_loss <- loss
  }
  
  rownames(p) <- as.character(unique_users)
  rownames(q) <- as.character(unique_movies)
  list(
    mu = mu,
    b_i = as.data.frame(b_i_dt),
    b_u = as.data.frame(b_u_dt),
    b_g = as.data.frame(b_g_dt),
    p = p,
    q = q
  )
}


fit <- fit_als_with_latent(train_set)
mu <- fit$mu
b_u <- setNames(fit$b_u$a, fit$b_u$userId)
b_i <- setNames(fit$b_i$b, fit$b_i$movieId)
b_g <- setNames(fit$b_g$c, fit$b_g$genres)
#
pq <- rowSums(fit$p[as.character(test_set$userId), ] * 
                fit$q[as.character(test_set$movieId), ])
test_set$pq <- pq
resid <- with(test_set, rating - clamp(mu + 
                                         b_i[as.character(movieId)] + 
                                         b_u[as.character(userId)] + 
                                         b_g[genres] + 
                                         pq))
rmse(resid)
```

One issue I encountered in an earlier version of the model was that around
the 85th or so iteration, the RMSE would start growing exponentially until the
`delta` and `loss` values became so large that R crashed entirely. I added a 
check so that if that occurred, the model would exit rather than crash. The 
cause was how I was calculating the latent factors. I was calculating `p` and `q` 
using the data table's `userId` and `movieId` columns instead of the indexing by
user and movie, thus erroneously trying to calculate a latent effect per rating
instead of per user and per movie. Creating vectors of unique user IDs and 
unique movie IDs, and then indexing movie ratings by user and user who gave ratings
by movie, got me back on the right track.

The largest challenge was performance. Many earlier versions of the model used 
nested loops and fewer indexes and took days--even at one point, just short of 
two weeks--to run. The key improvements to get it to run in a more reasonable 
amount of time were to calculate in two passes (first effects based on known 
factors, then latent effects) and to do as much pre-work as possible to index 
data in order to limit how many lookups needed to happen within the latent 
factor calculations. I also stopped using `rows_update()`, which I'd relied on 
previously to update the data table values in place.


## Final Results

When tested against the 20% of data that had been held out throughout the model
development, the RMSE was around 0.84, below the 0.86 target.
```{r final_test}
mu <- fit$mu
b_u <- setNames(fit$b_u$a, fit$b_u$userId)
b_i <- setNames(fit$b_i$b, fit$b_i$movieId)
b_g <- setNames(fit$b_g$c, fit$b_g$genres)

pq <- rowSums(fit$p[as.character(final_holdout_test$userId), ] * 
                fit$q[as.character(final_holdout_test$movieId), ])
final_holdout_test$pq <- pq
resid <- with(final_holdout_test, rating - clamp(mu + 
                                                   b_i[as.character(movieId)] + 
                                                   b_u[as.character(userId)] + 
                                                   b_g[genres] + 
                                                   pq))
rmse(resid)
```

While the model takes several minutes to run, it's a vast improvement over the performance I was seeing earlier, and
seems acceptable given the size of the dataset and that it's being run on a fairly ordinary computer.

## Conclusion

The combination of user, movie, genre, and latent effects improved the RMSE from a baseline of 1.059728 to
0.841691. That RMSE is validated by testing against a completely separate hold out test set that was not
used for any part of the model development.

### Limitations
While I am quite happy with my model, it does have limitations that could be improved upon in future iterations.
Most notably, the model does not handle movies or users that were not in its original training set. I worked
around this by assigning 20% of ratings per user to the test set and ensuring that if a user or movie had too
few data points to be in both sets, it ended up in the training set. However, a more resilient model would account
for unknown data.

Likewise, I made some simplifications to the model that make it less useful if
given a differently sized training set. I used constant values for lambdas to
make the values easier to comprehend and test, but that means a much smaller 
training set would be heavily overregularized. To make the model more flexible,
it could be re-tuned using lambdas that are then multiplied by the total number
of rows, now that the general values for a 7 million row set are known. Similarly,
in earlier versions I attempted to set a minimum number of ratings to use for
calculating latent effects, as the `fit_recommender_model` in the `dslabs` 
package does. I removed this as it added complexity to the calculations with
little impact -- since my dataset was large enough, the minimum ratings excluded
very few, if any, rows. However, to support varying sizes of training data, this
could be added.

Finally, given the length of time needed for the code to run, I did not try K values larger than 24. It is possible
that including more latent effects might further improve the predictions.

### Future Work
There are other ideas that I did not explore here that could be interesting to look it in a future model:

- I tried each unique combination of genres as a separate genre, since "Comedy|Horror," for example, does seem to me to
  be a different experience from "Horror". But that also means that it's more difficult to tell whether "Horror" has any
  overall effect on ratings. Splitting the genres might yield different results
- After determining decade did not impact rating, I did not go back and test whether release year impacted rating. I did
  not expect that it would, but that could be validated
- While I tested known factors without latent effects, I did not do the reverse. My assumption was that the known factors
  should be included in the model, but a latent-effect-only model could also be worth trying

## Acknowledgements
While I cannot cite every source I looked at online during the length of this
project, there were a few resources that I consulted multiple times and to which
I am particularly indebted.

- [_Introduction to Data Science: Data Wrangling and Visualization with R_ by Rafael A. Irizarray](https://rafalab.dfci.harvard.edu/dsbook-part-1/)
- [_Introduction to Data Science: Statistics and Prediction Algorithms Through Case Studies_ by Rafael A. Irizarray](https://rafalab.dfci.harvard.edu/dsbook-part-2/)
- ["dslabs: Data Science Labs" R package, specifically the source code for `fit_recommender_model`](https://cran.r-project.org/package=dslabs)
- ["Claude" by Anthropic, Opus 4.6 and Sonnet 4.6 models](https://claude.ai)